The prevalence of child welfare system contact among American Indian / Alaska Native children, 2003 - 2016

### 2020-01-01 ms.rmd
final update before 12/31 report

ncands_inv<-read_csv("./data/ncands_first_inv.csv") %>% 
  filter(year>=2003, year<2017) %>% 
  filter(age<18) %>% 
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White")
ncands_sub<-read_csv("./data/ncands_first_sub.csv")%>% 
  filter(year>=2003, year<2017)%>% 
  filter(age<18) %>% 
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White")
afcars_fc<-read_csv("./data/afcars_first_fc.csv")%>% 
  filter(year>=2003, year<2017)%>% 
  filter(age<18) %>%   
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White")
afcars_tpr<-read_csv("./data/afcars_tpr.csv")%>% 
  filter(year>=2003, year<2017)%>% 
  filter(age<18) %>% 
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White") 
afcars_non_icwa<-read_csv("./data/afcars_non_icwa.csv")%>% 
  filter(year>=2003, year<2017)%>% 
  filter(age<18) %>% 
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White")
afcars_fc_inv<-read_csv("./data/afcars_fc_post_inv.csv")%>% 
  filter(year>=2003, year<2017)%>% 
  filter(age<18) %>% 
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White")
```

@@

This report examines how american indian children are processed through
us child welfare systems. it explores the likelihood that american indian children eventually end up in foster care, end up in foster care with non-indian foster parents, or see the rights of their parents terminated. I compare these pathways through the foster care system to the pathways experienced by white and children.

**FINDINGS HERE**

@@

American Indian / Alaska Native (AI/AN) children and families are far more likely to experience contact with the child welfare system than are their white peers *(CITES)*. *STATEMENT ON SIGNIFICANCE*

This report provides national estimates of the prevalence of a series of child welfare system contact outcomes for AI/AN children, in addition to estimating the likelihood that children transition to more serious child welfare system outcomes conditional on lower-level contacts, such as investigations.

I use the 2003 - 2016 AFCARS and NCANDS to calculate period lifetables for lifetime risk of investigation, substantiation, placement, placement into a non-ICWA compliant foster care settings, and termination of parental rights. Each of these tables provides an estimate of the cumulative risk of experiencing the event before age 18 if risk-levels observed during the reported period remain constant.

For the national population of both AI/AN and white children, I estimate the following probabilities of event occurence by age, and as a cumulative likelihood of event occurrence before a child's 18th birthday:

1. Screened-in maltreatment report
2. Confirmed or substantiated maltreatment report
3. Foster care placement
3. Placement in foster home that is not compliant with ICWA guidelines (for AI/AN children)
4. Termination of both parent's rights while in foster care

@@

1. Substantiation, conditional on screened-in report
2. Foster care placement, conditional on screened-in report
3. Foster care placement, conditional on substantiated report
4. ICWA non-compliant placement, conditional on any foster care placement
5. Parental rights termination, conditional on any foster care placmeent

@@

Life table methods adjust for the variable risk of an event and the varying size of a population over the life course by following a cohort of children over their life course to observe the incidence of an event. Given shifts in risk over time, and the lack of longitudinal data on children over a sufficientily long period of time, I use a period life table approach, that creates a synthetic cohort of 100,000 children, and simulates the risk of experiencing each event based on the observed incidence in the population. This method relies on two key assumptions: 1) that event risk remains stable at observed levels for each age-group as the cohort ages and 2) that mortality rates for each age-group remain stable at observed levels as the cohort ages.

@@

Marginal probabilities can be interpreted as either a) the likelihood that an individual will experience an event or b) the proportion of the population that will experience an event. In sections **N THROUGH P** I estimate the marginal probability of a series of child welfare events for the complete population of AI/AN and white children.

Conditional probabilities describe the likelihood of an event, conditional on knowing that another event occurred. A probability of event $A$ conditional on event $B$, $Pr(A|B)$ can be interpreted as the likelihood that an individual who experiences event B will experience event A. Alternatively, we can interprent this probability as the proportion of individuals who experienced B that also experienced A. Formally, the conditional probability of a event A conditional on event B is defined as the ratio of the joint probability of A and B divided by the marginal probability of event B:

@@

Figure \@ref(fig:inv) panel B displays the age-specific risk of a first screened-in maltreatment report for AI/AN and white children using pooled 2012 - 2016 data. Risk is highest in the first years of life, and risk for AI/AN children is markedly higher than risk for white children during infancy. AI/AN infants are 1.44 times more likely than white infants to recieve a screened-in maltreatment report. Between 2012 and 2016, 7.7 percent of AI/AN infants recieve a report, compared to 5.3 percent of white infants. AI/AN children are at elevated risk until age 6. At later ages, risk profiles for AI/AN and white children converge.

@@

Next, I estimate the marginal probability of ever experiencing a maltreatment investigation before age 18, and the probability of experiencing a first investigation at each age between birth and 17. Panel A in Figure \@ref(fig:sub) shows how the lifetime risk of substantiation has shifted over time. Unlike investigations, substantiation risk levels have not clearly increased or decreased across this time period. AI/AN risk levels fluctuated around the period-average level of 0.14, whereas white risk levels centered around 0.1. In 2016, AI/AN children's risk of substantiation before age 18 was 0.16, and white chilren's risk was 0.1.

@@

ggplot(ncands_sub_inv,
         aes(x = age, y = sub_inv, color = race_ethn)) +
  geom_line() + 
  geom_point() + 
  xlab("Age") + 
  ylab("Risk of event")
```

Figure \@ref(fig:subCond) shows the age-specific probability of a first maltreatment report substantiation, conditional on having a prior maltreatment report (inclusive of a first report). AI/AN children are substantially more likely than white children to have a substantiated maltreatment report than white children. Among AI/AN infants that have been investigated by a child welfare agency, about 41 percent will be substantiated. For white infants, about 36 percent that receive an investigation will be subtantiated. The gap is largest at age 9, when AI/AN children with a report are 1.6 times more likely than white children with a report to experience a substantiated report.

@@

```{r}
## P(FC,inv)
fc_post_inv_tables<-list()
years<-unique(afcars_fc_inv$year)
race_id<-unique(afcars_fc_inv$race_ethn)
index<-1
for(i in 1:length(unique(afcars_fc_inv$.imp))){
  for(r in 1:length(race_id)){
    temp<-afcars_fc_inv %>%
      filter(year>2011)
    temp<-temp %>% 
      filter(race_ethn == race_id[r])
    temp<-temp %>% 
      filter(.imp==i)
    temp<-temp %>% 
      group_by(.imp, age, race_ethn) %>% 
      summarise(var = sum(var), pop = sum(pop)) %>% 
      ungroup()
    fc_post_inv_tables[[index]]<-make_life_table(temp)
    index<-index+1
  }
}

# P(A,B)
fc_post_inv_tables<-bind_rows(fc_post_inv_tables)

@@

ggplot(fc_cond_inv,
       aes(x = age, y = fc_inv, 
           color = race_ethn)) + 
  geom_line()
```

@@

The Indian Child Welfare Act, among other protections including tribal jurisdiction over AI/AN children, guarantees native children in foster care the right to be in a placement that preserves their heritage *CITE MEASURING COMPLIANCE*. Preferences are guaranteed for kin, and non-kin members of the child's nation. Otherwise, AI/AN foster families from outside of the child's tribe or nation are to be preferred. Placement into a non-native non-kin placement is to be considered a last resort.

Figure \@ref(fig:icwa) panel A shows the lifetime risk of ever being placed into a non-kin non-native foster placement for AI/AN children over time. Note that AFCARS only records the final placement setting for each child in a reporting year, so these estimates are likely negatively biased. AI/AN children's lifetime risk of ever being placed in a non-kin non-native foster placement has declined over time, from a peak of 0.12 in 2005 to a low of 0.08 in 2016. Panel B displays the age-specific risk of non-kin, non-native foster placement for AI/AN children. As with other foster placements, risk is highest in infancy, and declines as children age. **CALL OUT CONDITIONAL PROB FIG**

@@

@@

```{r tprTS,  fig.pos="h", out.extra = '', fig.cap="Risk of termination of parental rights while in foster care before age 18, 2003 - 2016 period life table estimates"}
tpr_tables<-list()
years<-unique(afcars_tpr$year)
race_id<-unique(afcars_tpr$race_ethn)
index<-1
for(i in 1:length(unique(afcars_tpr$.imp))){
  for(j in 1:length(years)){
    for(r in 1:length(race_id)){
      temp<-afcars_tpr %>%
        filter(year==years[j])
      temp<-temp %>%
        filter(race_ethn == race_id[r])
      temp<-temp %>%
        filter(.imp==i)
      tpr_tables[[index]]<-make_life_table(temp)
      index<-index+1
    }
  }
}
tpr_tables<-bind_rows(tpr_tables)

@@

tpr_tables_c %>%
  ggplot(aes(x=year, y = c_mn,
             fill = race_ethn,
             group = race_ethn,
             ymin = cmin, ymax = cmax)) +
  geom_ribbon(alpha = .5) +
  geom_line(aes(color = race_ethn)) +
  xlab("Year") +
  ylab("Risk of event") + 
  labs(fill = "", color = "")
```

## P(FC|sub)
```{r}

# fc_cond_sub<-fc_post_inv_tables %>%
#   mutate(fc_sub = q / subst_tables$q)
# 
# ggplot(fc_cond_sub %>%
#          filter(year==2016),
#          aes(x = age, y = fc_sub, color = race_ethn)) +
#   geom_line()
```

## P(FC|tpr)
```{r}
### use bayes' rule because p(fc|tpr)=1
# \[P(B|A) = \frac{P(A|B)P(B)}{P(A)}\]
## P(tpr|fc) = p(fc|tpr)p(tpr) / p(fc)
## p(tpr|fc) = p(tpr)/p(fc)
# fc_tpr_fc<-tpr_tables %>%
#   mutate(tpr_fc = q/ fc_tables$q)
# 
# ggplot(fc_tpr_fc %>%
#          filter(year==2016),
#          aes(x = age, y = tpr_fc, color = race_ethn)) +
#   geom_line()

NEXT: compare these to raw rates,

Compare to lifetime risk of ever investigated

CONDITIONALS ONLY WORK FROM SAME DATA - CANT ASSUME THAT p(b|a)=1 otherwise

need to go back to original afcars - rely on imputed ncands to get uncertainty, link to afcars by id, check year for investigation incident, maybe +/-1 yr for lags in placement starts

want to use the proportions to obtain information about probability. perhaps use a model to populate uncertainty? or construct CI from data - either way should be ok

>>>>

ncands_inv<-read_csv("./data/ncands_first_inv.csv") %>% 
  filter(year>=2003, year<2017) %>% 
  filter(age<18) %>% 
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White")
ncands_sub<-read_csv("./data/ncands_first_sub.csv")%>% 
  filter(year>=2003, year<2017)%>% 
  filter(age<18) %>% 
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White")
afcars_fc<-read_csv("./data/afcars_first_fc.csv")%>% 
  filter(year>=2003, year<2017)%>% 
  filter(age<18) %>%   
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White")
afcars_tpr<-read_csv("./data/afcars_tpr.csv")%>% 
  filter(year>=2003, year<2017)%>% 
  filter(age<18) %>% 
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White") 
afcars_non_icwa<-read_csv("./data/afcars_non_icwa.csv")%>% 
  filter(year>=2003, year<2017)%>% 
  filter(age<18) %>% 
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White")
afcars_fc_inv<-read_csv("./data/afcars_fc_post_inv.csv")%>% 
  filter(year>=2003, year<2017)%>% 
  filter(age<18) %>% 
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White")
afcars_fc_sub<-read_csv("./data/afcars_fc_post_sub.csv")%>% 
  filter(year>=2003, year<2017)%>% 
  filter(age<18) %>% 
  filter(.imp!=0) %>% 
  filter(race_ethn == "AI/AN" | race_ethn == "White")
```

@@

This report examines how American Indian / Alaska Native children are processed through state and local child welfare systems. It establishes age-specific and lifetime prevalence of a series of child welfare system events, then evaluates how likely cases are to move "up" the chain of more serious outcomes, conditional on experiencing a lower-level outcome. In doing so, it shows that inequalities in child welfare system outcomes for AI/AN children emerge at distinct stages of case processing. These national data provide a high-level overview of system inequalities, but may mask important sub-national variation.

- *Investigation*: Nationally, AI/AN children only experience inequality in investigations at young ages: between birth and age 5. As they age, risk of investigation converges to the risk levels for white children. 
- *Substantiation*: However, AI/AN children face a higher risk of substantiation than white children at all ages. Conditional on receiving an investigation, AI/AN children are at much higher risk of substantiation than are white children. 
- *Foster care*: AI/AN children face much higher risk of foster care than white children at all ages, and these gaps persist conditional on maltreatment investigation or maltreatment substantiation. 
- *Placement in non-kin, non-native foster settings*: An overwhelming majority of AI/AN children who ever enter foster care experience placement in a setting that does not meet the standards for appropriate placements for AI/AN children specified in the Indian Child Welfare Act
- *Termination of parental rights*: AI/AN children are more likely than white children to have their parents' rights terminated through the child welfare system. However, this inequality appears to be a function of higher levels of foster care placement. Conditional on ever being placed in foster care, AI/AN risks are similar to white risks.

@@

American Indian / Alaska Native (AI/AN) children and families are far more likely to experience contact with the child welfare system than are their white peers [@haightScopingStudyIndigenous2018]. This report provides national estimates of the prevalence of a series of child welfare system contact outcomes for AI/AN children, in addition to estimating the likelihood that children transition to more serious child welfare system outcomes conditional on lower-level contacts, such as investigations.

I use the 2003 - 2016 AFCARS and NCANDS to calculate period life tables for lifetime risk of investigation, substantiation, placement, placement into a non-ICWA compliant foster care settings, and termination of parental rights. Each of these tables provides an estimate of the cumulative risk of experiencing the event before age 18 if risk-levels observed during the reported period remain constant.

For the national population of both AI/AN and white children, I estimate the following probabilities of event occurrence by age, and as a cumulative likelihood of event occurrence before a child's 18th birthday:

1. Screened-in maltreatment report
2. Confirmed or substantiated maltreatment report
3. Foster care placement
3. Placement in a non-kin, non-AI/AN foster home 
4. Termination of both parent's rights while in foster care

@@

1. Substantiation, conditional on screened-in report
2. Foster care placement, conditional on screened-in report
3. Foster care placement, conditional on substantiated report
4. Placement in a non-kin, non-AI/AN foster home, conditional on any foster care placement
5. Parental rights termination, conditional on any foster care placement

@@

Life table methods adjust for the variable risk of an event and the varying size of a population over the life course by following a cohort of children over their life course to observe the incidence of an event. Given shifts in risk over time, and the lack of longitudinal data on children over a sufficiently long period of time, I use a period life table approach, that creates a synthetic cohort of 100,000 children, and simulates the risk of experiencing each event based on the observed incidence in the population. This method relies on two key assumptions: 1) that event risk remains stable at observed levels for each age-group as the cohort ages and 2) that mortality rates for each age-group remain stable at observed levels as the cohort ages.

@@

Marginal probabilities can be interpreted as either a) the likelihood that an individual will experience an event or b) the proportion of the population that will experience an event. I estimate the marginal probability of a series of child welfare events for the complete population of AI/AN and white children.

Conditional probabilities describe the likelihood of an event, conditional on knowing that another event occurred. A probability of event $A$ conditional on event $B$, $Pr(A|B)$ can be interpreted as the likelihood that an individual who experiences event B will experience event A. Alternatively, we can interpret this probability as the proportion of individuals who experienced B that also experienced A. Formally, the conditional probability of a event A conditional on event B is defined as the ratio of the joint probability of A and B divided by the marginal probability of event B:

@@

Figure \@ref(fig:inv) panel B displays the age-specific risk of a first screened-in maltreatment report for AI/AN and white children using pooled 2012 - 2016 data. Risk is highest in the first years of life, and risk for AI/AN children is markedly higher than risk for white children during infancy. AI/AN infants are 1.44 times more likely than white infants to receive a screened-in maltreatment report. Between 2012 and 2016, 7.7 percent of AI/AN infants receive a report, compared to 5.3 percent of white infants. AI/AN children are at elevated risk until age 6. At later ages, risk profiles for AI/AN and white children converge.

@@

Next, I estimate the marginal probability of ever experiencing a maltreatment investigation before age 18, and the probability of experiencing a first investigation at each age between birth and 17. Panel A in Figure \@ref(fig:sub) shows how the lifetime risk of substantiation has shifted over time. Unlike investigations, substantiation risk levels have not clearly increased or decreased across this time period. AI/AN risk levels fluctuated around the period-average level of 0.14, whereas white risk levels centered around 0.1. In 2016, AI/AN children's risk of substantiation before age 18 was 0.16, and white children's risk was 0.1.

@@

ggplot(ncands_sub_inv,
         aes(x = age, y = sub_inv, color = race_ethn)) +
  geom_line() + 
  geom_point() + 
  xlab("Age") + 
  ylab("Risk of event")+ 
  labs(fill = "", color = "")
```

## Substantiation, conditional on investigation

Figure \@ref(fig:subCond) shows the age-specific probability of a first maltreatment report substantiation, conditional on having a prior maltreatment report (inclusive of a first report). AI/AN children are substantially more likely than white children to have a substantiated maltreatment report than white children. Among AI/AN infants that have been investigated by a child welfare agency, about 41 percent will be substantiated. For white infants, about 36 percent that receive an investigation will be substantiated. The gap is largest at age 9, when AI/AN children with a report are 1.6 times more likely than white children with a report to experience a substantiated report.

@@

## Foster care, conditional on investigation and substantiation

AI/AN children are more likely to face more serious consequences of maltreatment investigations and confirmed maltreatment reports than their white peers. Conditional on an investigation, or conditional on a substantiation, AI/AN children are substantially more likely than white children to ever enter foster care. Figure \@ref(fig:fcCond) shows the conditional probabilities of ever entering foster care following a first maltreatment investigation (panel A) and a first confirmed maltreatment investigation (panel B) for each age using life tables estimated from 2012 - 2016 pooled data. About 26 percent of AI/AN infants who are investigated will later be placed into foster care, compared to about 15 percent of white infants.  About 54 percent of AI/AN infants with a substantiated report will ever be placed into foster care, compared to 38 percent of white infants.

```{r fcCond,  fig.pos="h", out.extra = '', fig.cap = "(A): Age-specific risk of first foster care entry conditional on screened-in maltreatment report, 2012 - 2016 pooled period life table. (B):  Age-specific risk of first foster care entry conditional on substantiated maltreatment report, 2012 - 2016 pooled period life table"}
## P(FC,inv)
fc_post_inv_tables<-list()
years<-unique(afcars_fc_inv$year)
race_id<-unique(afcars_fc_inv$race_ethn)
index<-1
#for(i in 1:length(unique(afcars_fc_inv$.imp))){
  for(r in 1:length(race_id)){
    temp<-afcars_fc_inv %>%
      filter(year>2011)
    temp<-temp %>% 
      filter(race_ethn == race_id[r])
    temp<-temp %>% 
      filter(.imp==1)
    temp<-temp %>% 
      group_by(.imp, age, race_ethn) %>% 
      summarise(var = sum(var), pop = sum(pop)) %>% 
      ungroup()
    fc_post_inv_tables[[index]]<-make_life_table(temp)
    index<-index+1
  }
#}
# P(A,B)
fc_post_inv_tables<-bind_rows(fc_post_inv_tables)

@@

##P(FC,sub)
fc_sub_tables<-list()
years<-unique(afcars_fc_sub$year)
race_id<-unique(afcars_fc_sub$race_ethn)
index<-1
#for(i in 1:length(unique(afcars_fc_sub$.imp))){
  for(r in 1:length(race_id)){
    temp<-afcars_fc_sub %>%
      filter(year>2011)
    temp<-temp %>% 
      filter(race_ethn == race_id[r])
    temp<-temp %>% 
      filter(.imp==1)
    temp<-temp %>% 
      group_by(.imp, age, race_ethn) %>% 
      summarise(var = sum(var), pop = sum(pop)) %>% 
      ungroup()
    fc_sub_tables[[index]]<-make_life_table(temp)
    index<-index+1
  }
#}
# P(A,B)
fc_sub_tables<-bind_rows(fc_sub_tables)
# P(A|B) = P(A,B)/P(B)
fc_sub_cond<-fc_sub_tables %>%
  mutate(fc_sub = q / sub_tables_pooled$q)

inv<-ggplot(fc_cond_inv,
       aes(x = age, y = fc_inv,
           color = race_ethn)) +
  geom_line() + 
  geom_point() + 
  xlab("Age") + 
  ylab("Risk of event") + 
  labs(fill = "", color = "",
       title = "(A)")

sub<-ggplot(fc_sub_cond,
            aes(x = age, y = fc_sub,
                color = race_ethn)) + 
  geom_line() + 
  geom_point() + 
  xlab("Age") + 
  ylab("Risk of event") + 
  labs(fill = "", color = "",
       title = "(B)")

grid.arrange(inv, sub)

### NEXT DO p(fc,subst)
```

@@

The Indian Child Welfare Act, among other protections including tribal jurisdiction over AI/AN children, guarantees native children in foster care the right to be in a placement that preserves their heritage [@williamsIndianChildWelfare2015]. Preferences are guaranteed for kin, and non-kin members of the child's nation. Otherwise, AI/AN foster families from outside of the child's tribe or nation are to be preferred. Placement into a non-native non-kin placement is to be considered a last resort.

Figure \@ref(fig:icwa) panel A shows the lifetime risk of ever being placed into a non-kin non-native foster placement for AI/AN children over time. Note that AFCARS only records the final placement setting for each child in a reporting year, so these estimates are likely negatively biased. AI/AN children's lifetime risk of ever being placed in a non-kin non-native foster placement has declined over time, from a peak of 0.12 in 2005 to a low of 0.08 in 2016. Panel B displays the age-specific risk of non-kin, non-native foster placement for AI/AN children. As with other foster placements, risk is highest in infancy, and declines as children age. Note that these are population risks; risks only for those children in foster care are presented in Figure \@ref(fig:icwaCond).

@@

## Placement outside of kin and/or native foster homes conditional on foster care entry

Conditional on a foster care entry, an overwhelming majority of AI/AN children experience a foster care placement in a non-kin setting where the primary caretaker is not American Indian / Alaska Native. Using a 2012 - 2016 pooled life table, Figure \@ref(fig:icwaCond) shows the probability of being placed in a non-kin foster setting with neither primary caretaker being reported as AI/AN, conditional on a first foster care placement at each age. At no age is this risk less than 0.7. Young children have the highest probability of being placed in placement settings specified as preferred under ICWA. When children experience their first placement as teenagers, their risk of being placed in a non-kin non-AI/AN setting increases dramatically, to a peak of 0.95 at age 15.

```{r icwaCond,  fig.pos="h", out.extra = '', fig.cap = "AI/AN Age-specific risk of being placed in non-kin non-AI/AN caretaker foster care conditional on foster care entry, 2012 - 2016 pooled period life table"}
## P(non-icwa|placement)
## P(non-icwa, placement)/P(placement)
fc_tables_pooled_aian<-fc_tables_pooled %>% 
  filter(race_ethn=="AI/AN")

icwa_cond<-icwa_tables_pooled %>% 
  mutate(icwa_cond = q / fc_tables_pooled_aian$q)

ggplot(icwa_cond,
         aes(x = age, y = icwa_cond)) +
  geom_line() + 
  geom_point() + 
  xlab("Age") + 
  ylab("Risk of event") 
```

@@

AI/AN children are more likely than white children to experience the termination of both parents' rights while in foster care than are white children. Figure \@ref(fig:tpr) shows shifts in cumulative risk over time in panel A, and age-specific risks in recent years in panel B. Note the decline in probability in 2015 and 2016 is in part, a function of children entering foster care in those years having a shorter duration of exposure to risk; additional years of data collection may reveal upward revisions in those estimates. For all years included in this analysis, AI/AN children were at higher risk of having their parents' rights terminated. In the average year during this period (excluding 2015 and 2016), about 3 percent of all AI/AN children had their parents' rights terminated through the foster care system, compared to a risk of about 1 percent for white children.

Panel B shows the age-specific probability of termination of parental rights while in foster care for the full child population using 2012 - 2016 pooled data for life table estimates. As with other foster care outcomes, infants are at highest risk. AI/AN children are at higher risk than white children. From birth to age 1, AI/AN infants face a risk 2.3 times higher than white infant risk of parental rights termination.

```{r tpr,   fig.pos="h", out.extra = '', fig.cap = "(A): Risk of termination of parental rights while in foster care by age 18, 2003 - 2016 annual period life table. (B): Age-specific risk of termination of parental rights while in foster care, 2012 - 2016 pooled period life table"}
tpr_tables<-list()
years<-unique(afcars_tpr$year)
race_id<-unique(afcars_tpr$race_ethn)
index<-1
for(i in 1:length(unique(afcars_tpr$.imp))){
  for(j in 1:length(years)){
    for(r in 1:length(race_id)){
      temp<-afcars_tpr %>%
        filter(year==years[j])
      temp<-temp %>%
        filter(race_ethn == race_id[r])
      temp<-temp %>%
        filter(.imp==i)
      tpr_tables[[index]]<-make_life_table(temp)
      index<-index+1
    }
  }
}
tpr_tables<-bind_rows(tpr_tables)

tpr_tables_pooled<-list()
race_id<-unique(afcars_tpr$race_ethn)
imps<-unique(afcars_tpr$.imp)
index<-1
for(i in 1:imps){
  for(r in 1:length(race_id)){
    temp<-afcars_tpr %>% 
      filter(year>2011) 
    temp<-temp %>% 
      filter(race_ethn == race_id[r])
    temp<-temp %>% 
      filter(.imp==i)
    temp<-temp %>% 
      group_by(.imp, age, race_ethn) %>% 
      summarise(var = sum(var), pop = sum(pop)) %>% 
      ungroup()
    tpr_tables_pooled[[index]]<-make_life_table(temp)
    index<-index+1
  }
}
tpr_tables_pooled<-bind_rows(tpr_tables_pooled)

@@

ts<-tpr_tables_c %>%
  ggplot(aes(x=year, y = c_mn,
             color = race_ethn,
             group = race_ethn,
             ymin = cmin, ymax = cmax)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Risk of event") +
  labs(fill = "", color = "",
       title = "(A)") + 
  theme(plot.title = element_text(hjust = 0.5))

age<-tpr_tables_pooled %>% 
  ggplot(aes(x=age, y = q,
             color = race_ethn)) +
  geom_line() + 
  geom_point() + 
  xlab("Age") + 
  ylab("Risk of event") + 
  labs(fill = "", color = "",
       title = "(B)") + 
  theme(plot.title = element_text(hjust = 0.5))

Figure \@ref(fig:tprCond) shows the probability of parental rights termination, conditional on ever entering foster care. Unlike other child welfare system outcomes examined here, AI/AN children's risk of having parental rights termination, conditional on ever entering foster care is similar to or slightly lower than white children's risk at all ages. About 36 percent of AI/AN children who entered foster care during infancy saw their parents' rights later terminated, compared to about 40 percent of white children who entered foster care during infancy.

```{r tprCond,  fig.pos="h", out.extra = '', fig.cap = "Age-specific risk of termination of parental rights while in foster care conditional on foster care entry, 2012 - 2016 pooled period life table"}
## conditionals
tpr_cond<-tpr_tables_pooled %>% 
  mutate(tpr_cond = q / fc_tables_pooled$q)

ggplot(tpr_cond,
         aes(x = age, y = tpr_cond,
             color = race_ethn)) +
  geom_line() + 
  geom_point() + 
  xlab("Age") + 
  ylab("Risk of event") + 
  labs(fill = "", color = "")
```

# Conclusion

This report provides clear evidence that AI/AN children face unequal outcomes throughout non-tribal child welfare systems at a national level. It shows that the investigation stage (for young children), the substantiation decision stage, and the placement decision stage all contribute to inequalities. At the national level, parental rights termination decisions for children in foster care are similar for white and AI/AN children.

These descriptive findings cannot attribute causation to any stage of child welfare decision making, however. It is likely that incident-level, child-level, family-level, community-level, agency-level, and state-level characteristics all impact the likelihood that a child becomes subject to a report, and experiences particular trajectories as they are processed through a child welfare system. Future research should be closely attentive to how characteristics of the reported maltreatment episode and contextual characteristics, such as bias, law, and resources impact case outcomes.

These findings must be interpreted in the context of the legacy of the child welfare system in AI/AN communities. Historically, family separation and fostering have been used as a tool of forced assimilation and cultural genocide [@jacobsGenerationRemovedFostering2014; @crossAmericanIndianEducation1999; @haightScopingStudyIndigenous2018]. These findings show that despite landmark federal legislation and substantial efforts to build autonomous institutions for tribal governments to independently operate child welfare systems [@tropeSurveyAnalysisSelect2014], family separation through child welfare systems is still a common experience for AI/AN families.

# References {-#references}
